---
title: "Crash Course in R"
author: "Matthew E. Aiello-Lammens"
output: pdf_document
---


## Use a script file

Let's all make a new directory to work in today. 

Let's create a new script file in that directory.

Scripts make it easier to repeat your work. You can also add comments using the pound sign.

Short cut to execute commands and functions:

Mac:
[Command] + return

Windows
[Control] + return (RStudio)
[Control] + r (R gui)


### Setting your working directory

```{r, eval=FALSE}
setwd("[your dir name here]")
```


## R as a calculator

```{r}
5 + 3

8^3

6*3-1

# Order of operations follows PEMDAS
6*(3-1)
```

## Variables

We can save things within our session as **variables**

```{r}
pop_1 <- 1200
pop_2 <- 500

pop_total <- pop_1 + pop_2

pop_1 * 2

# Etc.
```

### Challange

I chagne pop_1

```{r}
pop_1 <- 2000
```

What is `pop_total` now?


## Loading Data

Best to use the full path to your data, but could also change into the directory you're data is in, then call it in there.


```{r}
bottlenose <- read.csv(file = "~/Dropbox/Projects/SCCC-Stats-Workshop/Symons_data.csv")
```


Let's have a look at these data

```{r}
head(bottlenose)

tail(bottlenose)

summary(bottlenose)

names(bottlenose)

str(bottlenose)
```


## Subsetting data

Let's work with only a subset of this data, selecting from the `data.frame` by columns.

```{r}
bottlenose_subset <- bottlenose[c("divetime", "logdivetime", "bouts")]
```

What if we wanted to subset by row/column number?

```{r}
bottlenose_subset[1, 1]
bottlenose_subset[5, 2]
```

Let's get a specific row.

```{r}
bottlenose_subset[3, ]
```

And a whole column

```{r, results='hide'}
bottlenose_subset[ , 2 ]
```

Just part of the column

```{r}
bottlenose_subset[1:10, 2 ]
```


Specific rows

```{r}
bottlenose_subset[c(3, 5, 7), ]
```


## Simple calculations / built-in functions

Some statistics of note.

```{r}
mean(bottlenose_subset$divetime)
max(bottlenose_subset$divetime)
median(bottlenose_subset$divetime)
```

What about getting this for all three columns at once?

```{r}
apply(X = bottlenose_subset, MARGIN = 2, FUN = mean)
```

We could also do this by row.

```{r, results='hide'}
apply(X = bottlenose_subset, MARGIN = 1, FUN = mean)
```

But we should be careful, because in this case, it doesn't make any sense to do this.

### Challenge

Use indexing and the functions we just learned to determin the mean, min, and max of rows 21 to 45.


## Simple plots

```{r}
plot(x = bottlenose_subset$bouts, y = bottlenose_subset$divetime)
```


### Challenge

Plot only the values from rows 21 to 45.