---
title: "Demographic Modeling Introduction"
author: "Matthew E. Aiello-Lammens"
output: 
  html_document: 
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Demographic models and population viability analysis (PVA)



## Calculating a population growth rate

Use census data from multi-year bottlenose dolphin census to calculate annual growth rates.

```{r bobblenose_census, eval=FALSE}
R <- rnorm(n = 25, mean = 1.05, sd = .01)

n_init <- 1000
n <- vector(length = length(R))

n_old <- n_init
for (t in 1:length(R)){
  n_new <- n_old * R[t]
  n[t] <- n_new
  n_old <- n_new
}

n <- floor(n)
n <- c(n_init, n)

bottlenose_census <- data.frame( year = 1990:2015,
                                 pop_size = n )

write.csv(file = "bottlenose_census.csv", x = bottlenose_census)

```



## Simulating population growth

### Bootstrap method



### Statistical distribution methods



## Calculating extinction risk and other measures



## Considering other sources of variability


## Caveats regarding PVA

* Be careful
* Only as good as data
* "All models are wrong, but some are useful" George Box


## Matrix projection models


